Random Forests

### 2020-07-01 pkg/partykit/inst/ULGcourse-2020/slides_forest.Rnw
revisions on section 'aggregation & sampling'; added example (wages)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3030 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 60f7cbeb1753c18ab5d34125d06830ac9e22d2dd

library("disttree")
library("Formula")
library("latex2exp")
library("lattice")
library("MASS")
library(colorspace)

@@

{\bf Idea:} \\
\begin{itemize}
\item Combination of an ensemble of trees. 
\item Each tree can capture non-linear and non-additive effects and select covariates and possible interactions automatically.
\item Combining trees to a forest model allows for an approximation of smooth effects.
\item A forest can regularize and stabilize the model.
\end{itemize}

@@

\subsection{Aggregation of trees\\
\&\\
supsampling}

@@

The predicted response $\hat{y}$ 
%or, more generally, the predicted model parameter (vector) $\hat{\theta}$ 
is the argument %of the parameter space $\Theta$ 
that minimizes the weighted sum of the objective function $f$ evaluated for learning data $\{y_i\}_{i=1,\ldots,n}$:

@@

\[
\hat{y}(x) =  \argmin_{y} \sum_{i=1}^n w_i(x) \cdot f(y; y_i)
\]

@@

\only<5->{
%\vspace{-3cm}
{\bf Note:} In model-based trees and forests a prediction is made not only for the response but for a model parameter (vector) $\theta$. 
This more general approach holds for classic random forests averaging over mean values of tree nodes by setting $f(\theta, y_i) = (\theta - y_i)^2$\\
but also allows for combining quantiles (quantile regression forests) or full distributions as in distributional forests with
$f(\theta, y_i) = log(likelihood(\theta, y_i))$.\\
}

@@

\begin{frame}
\frametitle{Subsampling}
{\bf Idea:} Build each tree on a different subset of the full data set.

@@

{\bf Types of subsampling:}
\begin{itemize}
\item Subsets (recommended size: 63.2\% of the full data set)
\item Bootstrapping (resampling with replacement)
\end{itemize}

@@

{\bf Random input variable sampling:} In order to reduce the correlation among trees only a subset of possible split variables is provided for each split (recommended size: $\sqrt{\text{\#Var}}$).

@@

{\bf Note:} Each tree is built as large as possible which can lead to overfitting for the single tree. However, this is compensated by aggregating the trees to a forest model.

@@

\subsection{Software}

@@

<<call_ranger, eval=FALSE, echo=TRUE>>=
ranger(formula, data)
@

@@

<<call_cforest, eval = FALSE>>=
cforest(formula, data, subset, ...)
@

@@

<<call_predict, eval=FALSE, echo=TRUE>>=
predict(object, newdata, type = "response", ...)
@

@@

\subsection{Variable importance}

Depending on the fitted model the corresponding objective function can be employed as \emph{risk function} to measure accuracy by calculating importance scores (e.g., log-likelihood, nr. of misclassifications, \ldots).

{\bf Implementations:}
\begin{itemize}
\item \fct{importance} in the R package randomForest.
\item \fct{importance.ranger} in the R package ranger.
\item \fct{varimp} in the R package partykit.
\end{itemize}

%\subsection{References}

>>>>

library("partykit")
library("Formula")
library("latex2exp")
library("lattice")
library("MASS")
library("colorspace")

@@

{\bf Idea:} \\
\begin{itemize}
\item Combine an ensemble of trees. 
\item Each tree can capture non-linear and non-additive effects and select covariates and possible interactions automatically.
\item Combining trees to a forest model allows for an approximation of smooth effects.
\item A forest can regularize and stabilize the model.
\end{itemize}

@@

\subsection{Aggregation of trees\\
\&\\
random sampling}

\begin{frame}
\frametitle{Aggregation of trees}
{\bf Simple tree model:} Obtain predictions for the response $y$ by averaging over observations in the same terminal node $\mathcal{B}_x$: 
$$
\hat{y}= \frac{1}{|\mathcal{B}_x|} \sum_{i: y_i \in \mathcal{B}_x} y_i
\visible<2->{
\ =\ \sum_{i=1}^n w_i(x) \cdot y_i$$
}

{\bf Model-based tree:} Obtain predictions for model parameter(s) $\theta$ by optimizing an objective function $f$ for all observations in the same terminal node $\mathcal{B}_x$:
$$
\hat{\theta}= \argmin_{\theta} \sum_{i: y_i \in \mathcal{B}_x} f(\theta; y_i)
\visible<2->{
\ =\ \argmin_{\theta} \sum_{i=1}^n w_i(x) \cdot f(\theta; y_i)$$
}

{\bf Note:} The model-based approach allows for fitting complex models but also includes the specific case of averaging over all observations which corresponds to minimizing the residual sum of squares $\sum (Y - \hat{Y})^2$.

@@

The predicted model parameter (vector) $\hat{\theta}$ 
is the argument of the parameter space $\Theta$ 
that minimizes the weighted sum of the objective function $f$ evaluated for learning data $\{y_i\}_{i=1,\ldots,n}$:

@@

\[
\hat{\theta}(x) =  \argmin_{\theta} \sum_{i=1}^n w_i(x) \cdot f(\theta; y_i)
\]

@@

@@

\begin{frame}
\frametitle{Random sampling}
{\bf Idea:} Build each tree on a different subset of the full data set.

@@

{\bf Sampling observations:}
\begin{itemize}
\item Bootstrapping: resampling with replacement, usually ``n out of n''.
\item Subsampling: draw a subset without replacement (typically of the size of 63.2\% of the full data set, but can be a smaller fraction for large data sets).
\end{itemize}

@@

{\bf Sampling covariates:}\\
\smallskip
In order to reduce the correlation among trees random input variable sampling is employed, i.e., for each split only a subset of covariates is selected randomly and provided as possible split variables (typical sampling size: $\sqrt{\text{\#Var}}$).
\end{frame}

\begin{frame}
\frametitle{Hyperparameter selection}

Apart from the size and type of sampling two further control parameters have to be set for a forest model: 
\begin{itemize}
\item {\bf Number of trees:} A high number of trees allows for a more stable model and for a better approximation of smooth effects, however, at the cost of higher computational effort.
\item {\bf Size of the trees:} The size of each single tree can be controlled by setting stopping criteria such as a minimal number of observations in a node, maximum tree depth or a significance level if a statistical test is employed in the tree algorithm. However, in a forest model each tree is usually built as large as possible which can lead to overfitting for a single tree, but this is compensated by aggregating the trees to a forest model.
\end{itemize}
\end{frame}

<<data_wages, eval=TRUE, echo=FALSE, results=hide>>=
data("CPS1985", package = "AER")
@

\begin{frame}[fragile]
\frametitle{Example: wages}

\textbf{Data:} Random sample from the May 1985 US Current Population Survey.
A data frame containing \Sexpr{nrow(CPS1985)} observations on \Sexpr{ncol(CPS1985)} variables.

@@

\begin{tabular}{ll}
\hline 
Variable & Description\\
\hline 
\code{wage}       & Wage (in US dollars per hour). \\
\code{education}  & Education (in years).  \\
\code{experience} & Potential work experience (in years, \code{age - education - 6}). \\
\code{age}        & Age (in years). \\
\code{ethnicity}  & Factor: Caucasian, Hispanic, Other. \\
\code{gender}     & Factor: Male, Female. \\
\code{union}      & Factor. Does the individual work on a union job? \\
\hline 
\end{tabular}

\textbf{Model formula:} \code{log(wage) ~ education + experience + age + ethnicity + gender + union}.

@@

<<example_wages, eval=TRUE, echo=FALSE, results=hide>>=
set.seed(4)
f <- log(wage) ~ education + experience + age + ethnicity + gender + union
ct_cps <- ctree(f, data = CPS1985, alpha = 0.01)
cf <- cforest(f, data = CPS1985, ntree = 5) #, control =ctree_control(maxdepth = 5))
vimp <- varimp(cf , risk = "loglik")
@

\begin{frame}[fragile]
\frametitle{Example: wages}
Single tree model:
\setkeys{Gin}{width=0.99\textwidth}
<<example_wages_ctree, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
plot(ct_cps)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{Example: wages}
Forest model - tree I:
\setkeys{Gin}{width=0.99\textwidth}
<<example_wages_cforest1, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
#par(mfrow=c(2,2))
plot(party(cf[[1]][[1]], data = CPS1985), drop_terminal = TRUE)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{Example: wages}
Forest model - tree II:
\setkeys{Gin}{width=0.99\textwidth}
<<example_wages_cforest2, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
#par(mfrow=c(2,2))
plot(party(cf[[1]][[2]], data = CPS1985), drop_terminal = TRUE)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{Example: wages}
Forest model - tree III:
\setkeys{Gin}{width=0.99\textwidth}
<<example_wages_cforest3, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
#par(mfrow=c(2,2))
plot(party(cf[[1]][[3]], data = CPS1985), drop_terminal = TRUE)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{Example: wages}
Forest model - tree IV:
\setkeys{Gin}{width=0.99\textwidth}
<<example_wages_cforest4, eval=TRUE, echo=FALSE, fig=TRUE, width=13>>=
#par(mfrow=c(2,2))
plot(party(cf[[1]][[4]], data = CPS1985), drop_terminal = TRUE)
@
\end{frame}

\begin{frame}[fragile]
\frametitle{Example: wages}
\vspace{-0.5cm}
\setkeys{Gin}{width=0.94\textwidth}
<<example_wages_forest_prediction, eval=TRUE, echo=FALSE, fig=TRUE, width = 10>>=
nd <- data.frame(education = 13.02, #c(2:18),
                 experience = 17.82,
                 age = c((18:64)/1), #36.83,
                 ethnicity = "cauc",
                 region = "other",
                 gender = "male",
                 occupation = "worker",
                 sector = "other",
                 union = "no",
                 married = "yes")

{
  par(mfrow = c(2,2))
  set.seed(74)
  cf <- cforest(f, data = CPS1985, ntree = 2)
  predwage <- cbind(c((18:64)/1), predict(cf, newdata = nd))
  colnames(predwage) <- c("age", "predicted log(wage)")
  plot(x = predwage[,"age"], y = predwage[,"predicted log(wage)"], type = 'l',
       ylab = "predicted log(wage)", xlab = "age", main = "2 trees")

  cf <- cforest(f, data = CPS1985, ntree = 4)
  predwage <- cbind(c((18:64)/1), predict(cf, newdata = nd))
  colnames(predwage) <- c("age", "predicted log(wage)")
  plot(x = predwage[,"age"], y = predwage[,"predicted log(wage)"], type = 'l',
       ylab = "predicted log(wage)", xlab = "age", main = "4 trees")

  cf <- cforest(f, data = CPS1985, ntree = 20)
  predwage <- cbind(c((18:64)/1), predict(cf, newdata = nd))
  colnames(predwage) <- c("age", "predicted log(wage)")
  plot(x = predwage[,"age"], y = predwage[,"predicted log(wage)"], type = 'l',
       ylab = "predicted log(wage)", xlab = "age", main = "20 trees")

  cf <- cforest(f, data = CPS1985, ntree = 200)
  predwage <- cbind(c((18:64)/1), predict(cf, newdata = nd))
  colnames(predwage) <- c("age", "predicted log(wage)")
  plot(x = predwage[,"age"], y = predwage[,"predicted log(wage)"], type = 'l',
       ylab = "predicted log(wage)", xlab = "age", main = "200 trees")
}
@
\end{frame}

% {\bf Implementations:}
% \begin{itemize}
% \item \fct{importance} in the R package randomForest.
% \item \fct{importance.ranger} in the R package ranger.
% \item \fct{varimp} in the R package partykit.
% \end{itemize}

% \only<2->{
% \setkeys{Gin}{width=0.5\textwidth}
% <<example_wages_vimp, eval=TRUE, echo=FALSE, fig=TRUE, width=6, height=3>>=
% par(mar = c(4,5.5,0,2))
% barplot(sort(vimp, decreasing = FALSE),
%         horiz = TRUE, las = 1, axes = FALSE,
%         xlab = "Variable importance: mean decrease in log-likelihood")
% axis(1, at = seq(0,1,0.02), las = 1, mgp=c(0,1,0))
% @

Depending on the fitted model the corresponding objective function can be employed as \emph{risk function} to measure accuracy by calculating importance scores (e.g., log-likelihood, nr. of misclassifications, \ldots).
%}
\end{frame}

\begin{frame}
\frametitle{Example: wages}
Evaluation of variable importance in the fitted forest model:

\setkeys{Gin}{width=0.57\textwidth}
<<example_wages_vimp, eval=TRUE, echo=FALSE, fig=TRUE, width=6, height=3>>=
par(mar = c(4,5.4,1,2))
barplot(sort(vimp, decreasing = FALSE),
        horiz = TRUE, las = 1, axes = FALSE,
        xlab = "Variable importance: mean decrease in log-likelihood")
axis(1, at = seq(0,1,0.02), las = 1, mgp=c(0,1,0))
@
\end{frame}

\subsection{Implementations}

@@

<<call_ranger, eval=FALSE, echo=TRUE>>=
rf <- ranger(log(wage) ~ education + experience + age + ethnicity + gender + union,
             data = CPS1985)
@

@@

<<call_cforest, eval = FALSE>>=
cf <- cforest(log(wage) ~ education + experience + age + ethnicity + gender + union,
              data = CPS1985)
@

@@

For a ranger object:
<<call_predict_rf, eval=FALSE, echo=TRUE>>=
predict(rf, data = testdata)
@

For a cforest object:
<<call_predict_cf, eval=FALSE, echo=TRUE>>=
predict(cf, newdata = testdata)
@

@@

\subsection{References}
